Equalizer Programming Guide\\ {\footnotesize\mdseries \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf} {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}

### 2007-09-28 Developer/ProgrammingGuide/paper.tex
PG: update to v0.8
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\title{Equalizer Programming Guide\\
  {\footnotesize\mdseries
    \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
    {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}
}
\author{Stefan Eilemann\thanks{eile@eyescale.ch}\\[\medskipamount]
  % Eyescale Software Sarl
}
\date{
  \textbf{INCOMPLETE}\\[\medskipamount]
  Version 0.7, \today
}

\newcommand{\tm}{\texttrademark~}
\newcommand{\rc}{\raise 1ex\hbox{{\tiny\textregistered}}~}
\newcommand{\fig}[1]{Figure~\ref{#1}}
\newcommand{\link}[1]{\htmladdnormallink{#1}{#1}}

@@

\thispagestyle{empty}
\begin{figure}[ht]
  \hfill
  \includegraphics[width=4cm]{images/logo.pdf}\hfill
  \includegraphics[width=4cm]{images/vmml.pdf}\hfill\vspace{-1em}\\
%  {\small\htmladdnormallink{www.eyescale.ch}{http://www.eyescale.ch}}
\end{figure}
\vfill

@@

\clearpage
\tableofcontents
\thispagestyle{empty}
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    0.7     & Sep 21, 2007 & started channel and image compositing,
                             added event handling, moved to new eqPly
                             code base\\
    0.6     & Sep 14, 2007 & added pipe, window and object manager\\
    0.5     & Sep 3,  2007 & added config and partly pipe\\
    0.4     & Aug 31, 2007 & added distributed objects\\
    0.3     & Aug 26, 2007 & added application and render client\\
    0.2     & Aug 20, 2007 & added main function\\
    0.1     & Aug 19, 2007 & outlined the basic concepts\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

Equalizer provides a framework for the development of parallel
OpenGL\texttrademark\ 
applications. Equalizer-based applications can run a single
shared-memory system with multiple graphics cards (GPU's) or on a
distributed graphics cluster. This Programming Guide introduces the
programming interface using the \textsf{eqPly} example shipped with
Equalizer as a guideline.

@@

Equalizer is the next evolution of a generic parallel programming
interface for visualization applications using OpenGL. Existing
solutions, such as SGI's OpenGL Multipipe\texttrademark\ SDK, VRCO's
Cavelib\texttrademark\ and VRJuggler, implement a subset of concepts
similar to Equalizer. In other areas, e.g., tracking device support,
they provide more functionality.

Equalizer implements the minimum necessary layer to build parallel and
scalable OpenGL applications. The programmer structures the application
so that the OpenGL rendering can be executed in parallel, potentially
across multiple processes for cluster-based execution. Equalizer
provides the domain-specific parallel rendering know-how and abstracts
threading, synchronization, windowing and event handling for
portability. It is a `GLUT on steroids', providing parallel and
distributed execution, scalable rendering features and fully
customizable event handling.

@@

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/processes.pdf}
  {\caption{\small\label{fProcesses}Equalizer Processes}}
\end{wrapfigure}
All functionality of Equalizer is accessed through the Equalizer client
library, which implements all functionality discussed in this document.

@@

An Equalizer server is responsible for managing one visualization
system\footnote{a shared memory system or graphics cluster}. Currently
it is only useful for running one application at a time, but it will be
extended to support multiple applications concurrently and efficiently
on one system. The server controls and potentially launches the
application's rendering clients.

@@

The application connects to a server, which chooses a configuration for
the application. It provides a render client, to be launched by the
server. The application reacts on events and controls the rendering.

@@

The render client implements the rendering part of an application. It is
passive, and receives all its rendering tasks from the server. The tasks
are executed by calling the appropriate task methods (see
\ref{ssTaskMethods}).

The application might be a rendering client, in which case it can also
contribute to the rendering. It can choose not to implement any render
client-related code, in which case it is reduced to be the application's
'master' process without any OpenGL windows.

The rendering client can be the same executable as the application, as
is the case with \textsf{eqPly}. Real-world applications often implement
a separate, light-weight rendering client.

@@

Equalizer uses a C++ programming interface. The API is minimally
invasive, that is, Equalizer imposes only the minimal, natural execution
framework upon the application. It does not impose a scene graph or does
interfere in any way with the application's rendering code.

Methods called by the application always have the form
\textsf{verb[Noun]}, whereas methods called by Equalizer (`Task
Methods') have the form \textsf{nounVerb}.

@@

The root of the resource tree is the \textsf{eq::Config}, which
represents the current configuration of the application. It currently
only holds the local node, not all nodes of the configuration.

@@

An \textsf{eq::Node} is the representation of a single computer in the
system. It is one operating system process of the render client. All
node task methods are executed from the main application thread. Each
configuration might also use an application node, in which case the
application process is also used for rendering.

@@

The \textsf{eq::Pipe} is the abstraction of a graphics card (GPU). In
the current implementation it is also one operating system thread,
unless the pipe's thread hint is set to \textsf{false}. All pipe and
child window and channel task methods are executed from the pipe thread
for threaded pipes or from the main application thread for non-threaded
pipes\footnote{see also
  \link{http://www.equalizergraphics.com/documents/design/nonthreaded.html}}.

@@

An \textsf{eq::Window} is an drawable and OpenGL context. The drawable
can be an on-screen window or an off-screen PBuffer or
FBO\footnote{off-screen drawables are not yet implemented, but can be
  created by the application and used with Equalizer}.

@@

The \textsf{eq::Channel} is the abstraction of an OpenGL viewport within
its parent window. It is the entity executing the actual rendering.

\subsection{Resource Usage}

How the rendering resources are to be used is configured using a
compound tree. Each compound has a channel, which it uses to execute the
rendering tasks. The rendering tasks are computed by the server and send
to the render clients. At no point the application or render clients
have or need knowledge of compounds. The configuration of compounds is
not in the scope of this document\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/compounds.html}}.

@@

The \textsf{eqPly} example is shipped with the Equalizer distribution
and serves as a simple reference implementation of an Equalizer-based
application. Its focus is not on rendering features or visual quality.
It serves as a test bed for most of the Equalizer features.

@@

All classes in the example are in the \textsf{eqPly} namespace to avoid
type name ambiguities, in particular for the \textsf{Window} class.

@@

Versioned objects have to override \textsf{isStatic} to return false, in
order for Equalizer to know that they should be versioned. Right now,
only the master instance of the object is writable, that is,
\textsf{eqNet::Object::commit} can be called to generate a new
version.

@@

Equalizer supports multiple window system interfaces, at the moment
glX/X11, WGL and AGL/Carbon. Some operating systems, and therefore some
Equalizer versions, support multiple window systems\footnote{see
  \link{http://www.equalizergraphics.com/compatibility.html}}.

@@

@@

The central rendering routine is \textsf{Channel::frameDraw}. This
routine contains the application's OpenGL rendering code, which is being
rendered using the rendering context provided by Equalizer.

@@

The following contextual information has to be used in order to render
the view expected by Equalizer:
\begin{description}
\item[Buffer]: The OpenGL read and draw buffer as well as color mask.
  These parameters are influenced by the current eye pass, eye
  seperation and anaglyphic stereo settings.
\item[Viewport]: The two-dimensional pixel viewport restricting the
  rendering area within the channel. For correct operations, both
  \textsf{glViewport} and \textsf{glScissor} have to be used. The pixel
  viewport is influenced by the destination channel's viewport
  definition and compound viewports set for sort-first/2D decompositions.
\item[Frustum]: Frustum parameters as defined by
  \textsf{glFrustum}. The frustum is influenced by the destination
  channel's view definition, compound viewports, head matrix and the
  current eye pass.
\item[Head Transformation:] A transformation matrix positioning the
  frustum. This is typically an identity matrix and is used for off-axis
  frustra in immersive rendering.
\item[Range]: A one-dimensional range with the interval [0..1]. This
  parameter is optional and should be used by the application to render
  only the appropriate subset of its data.
\end{description}

@@

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/DB.pdf}
  {\caption{\small\label{fDB}Final view of an DB compound}}
\end{wrapfigure}

After the basic view setup, a directional light is set up, and the model
is positioned using the camera parameters from the frame data. The
camera parameters are transported using the the frame data to ensure
that all channels render the frame from the same position.

@@

Finally the model is rendered, if it could be loaded during
initialization. Otherwise a quad is drawn in its place:

@@

\if 0
{\footnotesize\begin{lstlisting}
void Channel::_drawModel( const Model* model )
{
    Window*                  window    = static_cast<Window*>( getWindow() );
    mesh::VertexBufferState& state     = window->getState();

@@

{\footnotesize\begin{lstlisting}
\end{lstlisting}}

The model data is organized a in 3-dimensional kD-tree\footnote{See also
  \link{http://en.wikipedia.org/wiki/Kd-tree}} for efficient view
frustum culling. During preprocessing, each node of the tree gets a
database range assigned.

{\footnotesize\begin{lstlisting}
\end{lstlisting}}
\fi

@@

This section discusses some features of interest which are not covered
by the \textsf{eqPly} section. Where possible, code examples from the
Equalizer distribution are used.

@@

Event handling is one of the areas which requires a lot of
flexibility. On one hand, the implementation differs slightly for each
window system due to conceptual differences in the OS-specific
implementation. On the other hand, each application and widget set has
its own model on how events are to be handled. Therefore, event handling
in Equalizer is customizable in any stage of the processing\footnote{\link{http://www.equalizergraphics.com/documents/design/eventHandling.html}}.

The default implementation aims to provide a quick and easy
implementation of event handling, while allowing all necessary
customizations. It gathers all events in the application main thread, so
that the developer only has to implement \textsf{Config::processEvent}
to update its data based on the pre-processed, generic events.

@@

Where possible, events are received and processed by a separate per-node
event thread to allow asynchronous event handling. Currently an event
thread is only used by the X11/glX window system. WGL receives and
processes the events from the pipe threads which created the
windows. AGL receives the events from application or node main
thread. Whenever the term \textbf{event thread} is used, it refers to
the thread receiving the event, i.e., a per-node thread for glX, the pipe
thread for WGL and the main thread for AGL.

@@

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/eventInit.pdf}
  {\caption{\small\label{fEventInit}Event Handling Initialization}}
\end{wrapfigure}

During window and pipe initialization the event handling is set up. For
both entities, \textsf{initEvent\-Handler} is called to register the pipe
or window with an event handler. This method may be overwritten to use a
custom event handler, or to not install an event handler to disable
event handling in Equalizer. Likewise, \textsf{exitEventHandler} is
called to de-initialize event handling.

An event handler consists of two parts: the generic base class providing
the interface and some generic functions, and the window-system-specific
part providiing the actual implementation.

@@

WGL and AGL require the application to manually receive and dispatch
(`pump') events. On WGL, this has to happen on each thread with windows,
whereas on AGL it has to happen only on the main thread. By default,
Equalizer pumps these events automatically for the application.

The methods \textsf{Client::useMessagePump} and
\textsf{Pipe::useMessagePump} can be overriden to return \textsf{false}
to disable this behaviour for their respective threads. On non-threaded
pipes, \textsf{Pipe::useMessagePump} is not called.

@@

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/eventHandling.pdf}
  {\caption{\small\label{fEventProcessing}Event Processing}}
\end{wrapfigure}

Events are received by an event handler. The event handler finds the
\textsf{eq::Window} associated to the event. It then creates a generic
\textsf{WindowEvent}, which holds important event data in a format
independent of the window system. The original event is attached to the
generic window event.

@@

Events sent using \textsf{Config::sendEvent} are queued in the
application thread. After a frame has been finished,
\textsf{Config::finishFrame} calls \textsf{Config::processEvents}. The
default implementation of this method provides non-blocking event
processing, that is, it calls \textsf{Config::processEvent} for each
queued event. By overriding this function, event-driven execution can be
implemented.

@@

\subsection{Image Compositing for Scalable Rendering}

@@

@@

A \textsf{eq::Frame} references a \textsf{eq::Fra\-me\-Data}. The frame data
is the object connecting output with input frames. It is a holder for
images and is used to signal the availability of the pixel data. An
\textsf{eq::Image} holds a two-dimensional snapshot of the framebuffer
and can contain color and/or depth information.

Readback and assemble operations on frames and images are designed to be
asynchronous. They have a start and finish method for readback and
assemble to allow the initiation and synchronization of the operation.
Currently, only synchronous readback and assembly using
\textsf{glReadPixels} and \textsf{glDrawPixels} is implemented in the
respective start method of the image.

\if 0

\begin{wrapfigure}{r}{.6\textwidth}
  \includegraphics[width=.6\textwidth]{images/assembly.pdf}
  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
\end{wrapfigure}

The offset of input and output frames characterizes the position of the
frame data with respect to the framebuffer, that is, the
\textbf{window's} lower-left corner. For output frames this is the
position of the channel in 
\fi

>>>>

\title{Equalizer Programming Guide\\
  {\footnotesize\mdseries
    \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
    {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}
}
\author{Stefan Eilemann\thanks{eile@eyescale.ch}\\[\medskipamount]
  % Eyescale Software Sarl
}
\date{
  \textbf{INCOMPLETE}\\[\medskipamount]
  Version 0.8, \today
}

\newcommand{\tm}{\texttrademark~}
\newcommand{\rc}{\raise 1ex\hbox{{\tiny\textregistered}}~}
\newcommand{\fig}[1]{Figure~\ref{#1}}
\newcommand{\sref}[1]{Section~\ref{#1}}
\newcommand{\link}[1]{\htmladdnormallink{#1}{#1}}

@@

\thispagestyle{empty}
\begin{figure}[ht]
  \center
  \includegraphics[width=4cm]{images/logo.pdf}
%  {\small\htmladdnormallink{www.eyescale.ch}{http://www.eyescale.ch}}
\end{figure}
\vfill
\begin{figure}[ht]
  \center
  \includegraphics[width=4cm]{images/vmml.pdf}
%  {\small\htmladdnormallink{www.eyescale.ch}{http://www.eyescale.ch}}
\end{figure}
\vfill

@@

\clearpage
\tableofcontents
\thispagestyle{empty}
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    0.8     & Sep 28, 2007 & added head tracking, finished channel,
                             proof-reading pass\\
    0.7     & Sep 21, 2007 & started channel and image compositing,
                             added event handling, moved to new eqPly
                             code base\\
    0.6     & Sep 14, 2007 & added pipe, window and object manager\\
    0.5     & Sep 3,  2007 & added config and partly pipe\\
    0.4     & Aug 31, 2007 & added distributed objects\\
    0.3     & Aug 26, 2007 & added application and render client\\
    0.2     & Aug 20, 2007 & added main function\\
    0.1     & Aug 19, 2007 & outlined the basic concepts\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

Equalizer provides a framework for the development of parallel
OpenGL\texttrademark\ applications. Equalizer-based applications can run
a single shared-memory system with multiple graphics cards (GPU's) or on
graphics clusters. This Programming Guide introduces the programming
interface using the \textsf{eqPly} example shipped with Equalizer as a
guideline.

@@

Equalizer is the next evolution of generic parallel programming
interfaces for visualization applications using OpenGL. Existing
solutions, such as SGI's OpenGL Multipipe\texttrademark\ SDK, VRCO's
Cavelib\texttrademark\ and VRJuggler, implement a subset of concepts
similar to Equalizer. In other areas, e.g., tracking device support,
they provide more functionality.

Equalizer implements the minimum necessary layer to build parallel and
scalable OpenGL applications. The programmer structures the application
so that the OpenGL rendering can be executed in parallel, potentially
using multiple processes for cluster-based execution. Equalizer provides
the domain-specific parallel rendering know-how and abstracts
configuration, threading, synchronization, windowing and event handling
for portability. It is a `GLUT on steroids', providing parallel and
distributed execution, scalable rendering features and fully
customizable event handling.

@@

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/processes.pdf}
  {\caption{\small\label{fProcesses}Equalizer Processes}}
\end{wrapfigure}
All functionality of Equalizer is accessed through the Equalizer client
library, which implements all functionality discussed in this
document. The client library provides communication between the Equalizer
processes.

@@

An Equalizer server is responsible for managing one visualization
system, i.e., a shared memory system or graphics cluster. Currently it
is only useful for running one application at a time, but it will be
extended to support multiple applications concurrently and efficiently
on one system. The server controls and potentially launches the
application's rendering clients.

@@

The application connects to an Equalizer server, which chooses a
configuration for the application. It also provides a render client, to
be launched by the server. The application reacts on events and controls
the rendering.

@@

The render client implements --obviously-- the rendering part of an
application. Its execution is passive, completely driven by rendering
tasks received from the server. The tasks are executed by calling the
appropriate task methods (see \sref{ssTaskMethods}) in the correct
thread and context.

The application might be a rendering client, in which case it can also
contribute to the rendering. If it does not implement any render
client-related code it is reduced to be the application's 'master'
process without any OpenGL windows.

The rendering client can be the same executable as the application, as
is the case with \textsf{eqPly}. When it is started as a render client,
the Equalizer initialization routine does not return and takes over the
control by calling the render client task methods. Complex real-world
applications often implement a separate, light-weight rendering client.

@@

Equalizer uses a C++ programming interface. The API is minimally
invasive, that is, Equalizer imposes only the minimal, natural execution
framework upon the application. It does not provide a scene graph or
does interfere in any other way with the application's rendering
code. The restructuring work needed for Equalizer is needed anyway in
order to parallelize the application for rendering.

Methods called by the application have the form \textsf{verb[Noun]},
whereas methods called by Equalizer (`Task Methods') have the form
\textsf{nounVerb}. For example the application calls
\textsf{Config::startFrame} to render a new frame, which causes --among
other things-- \textsf{Node::frameStart} to be called in all active
render clients.

@@

The root of the resource tree is the \textsf{eq::Config}, which
represents the current configuration of the application. The
configuration is the session in which all render clients are
registered. The config instance on a given node currently only holds the
local node, not all nodes of the configuration.

@@

An \textsf{eq::Node} is the representation of a single computer in the
system. One operating system process of the render client will be used
for each node. Each configuration might also use an application node, in
which case the application process is also used for rendering. All
node-specific task methods are executed from the main application
thread.

@@

The \textsf{eq::Pipe} is the abstraction of a graphics card (GPU). In
the current implementation it is also one operating system thread,
unless the pipe's thread hint is set to \textsf{false}. All pipe, window
and channel task methods are executed from the pipe thread for threaded
pipes or from the main application thread for non-threaded
pipes\footnote{see also
  \link{http://www.equalizergraphics.com/documents/design/nonthreaded.html}}.

@@

An \textsf{eq::Window} holds a drawable and OpenGL context. The
drawable can be an on-screen window or an off-screen PBuffer or
FBO\footnote{off-screen drawables are not yet implemented, but can be
  created by the application and used with Equalizer}. The window holds
window-system-specific handles to the drawable and context, e.g., an X11
window \textsf{XID} and \textsf{GLXContext} for the glX window system.

@@

The \textsf{eq::Channel} is the abstraction of an OpenGL viewport within
its parent window. It is the entity executing the actual rendering. The
channel's viewport might be overwritten when it is rendering for another
channel.

\subsection{Compounds}

How the rendering resources are to be used is configured using a
compound tree. Although the API does not currently expose compounds, the
basic design behind compounds is explained here to increase the
understanding of the architecture behind Equalizer.

Each compound has a channel, which it uses to execute the rendering
tasks. One channel might be used by multiple compounds. Unused channels
are not instantiated. The rendering tasks for the channels are computed
by the server and send to the appropriate render clients.

\begin{wrapfigure}{r}{.6\textwidth}
  \includegraphics[width=.6\textwidth]{images/frustra.pdf}
  {\caption{\small\label{fFrustra}Frustrum descriptions}}
\end{wrapfigure}
The channels of the leaf compounds (source channels) in the compound
tree execute the rendering for the top-most compound's channel
(destination channel). The top-most compound with a channel defines the
2D pixel viewport rendered by all leaf compounds. The destination
channel and pixel viewport can not be overridden.

Compounds have a frustum description, which is inherited to the
children. The frustum can be overridden by children, but this is rarely
used since there is no use case for this behaviour. The frustrum can be
specified as a wall, which is completely defined by the bottom-left,
bottom-right and top-left coordinates relative to the origin, or as a
projection, which is defined by the position and head-pitch-roll
orientation of the projector, as well as the horizontal and vertical
field-of-view and distance to the projection wall. \fig{fFrustra}
illustrates the wall and projection frustum parameters.

Compounds have attributes which configure the decomposition of the
destination's channel viewport, frustum and database. A
\textsf{viewport} decomposes the destination channel and frustum in
screen space. A \textsf{range} tells the application to render a part of
its database, and an eye pass can selectively render different stereo
passes.

Compounds have output and input frames which configure the recomposition
of the resulting pixels from the source compound channels. An output
frame connects to an input frame of the same name, and transports the
selected frame buffer data from the output channel to the input
channel. This composition process is extensively described in
\sref{sCompositing}.

More information on the configuration of compounds can be found on the
Equalizer website\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/compounds.html}}.

@@

The \textsf{eqPly} example is shipped with the Equalizer distribution
and serves as a reference implementation of an Equalizer-based
application. Its focus is not on rendering features or visual quality,
but it is used as a test bed for most of the Equalizer features.

@@

All classes in the example are in the \textsf{eqPly} namespace to avoid
type name ambiguities, in particular for the \textsf{Window}
class. \fig{fUml} shows the most important Equalizer classes and how
they are subclassed in the \textsf{eqPly} example. Each of the classes
in the \textsf{eqPly} namespace will be discussed in this section.

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/uml}
  {\caption{\small\label{fUml}UML diagram of significant Equalizer and
      eqPly classes and task methods.}}
\end{figure}

@@

Versioned objects have to override \textsf{isStatic} to return false, to
indicate that they are versioned. The current implementation has the
following characteristics:
\begin{itemize}
\item Only the master instance of the object is writable, that is,
  \textsf{eqNet::Object::com\-mit} can be called to generate a new
  version.
\item Slave instance versions can only be advanced, that is,
  \textsf{eqNet::Object::sync( version )} with a version smaller than
  the current version will fail.
\end{itemize}

@@

Equalizer supports multiple window system interfaces, at the moment
glX/X11, WGL and AGL/Carbon. Some operating systems, and therefore some
Equalizer versions, support multiple window systems
concurrently\footnote{see
  \link{http://www.equalizergraphics.com/compatibility.html}}.

@@

Parts of the Carbon API used for window and event handling in the AGL
window system are not thread safe. The application has to call
\textsf{eq::Global::enterCarbon} before any thread-unsafe Carbon call,
and \textsf{eq::Global::leaveCarbon} afterwards. These functions should
be used only during window initialization and exit, not during
rendering\footnote{for various reasons \textsf{enterCarbon} might block
  up to 50 milliseconds}. Carbon call in the window event handling
routine \textsf{Window::processEvent} are thread-safe. Please contact
the Equalizer developer mailing list if you need to use Carbon calls on
a per-frame basis.

@@

The central rendering routine is \textsf{Channel::frameDraw}. This
routine contains the application's OpenGL rendering code, which is being
rendered using the rendering context provided by Equalizer. As most of
the other task methods, \textsf{frameDraw} is called in parallel by
Equalizer on all pipe threads in the configuration. Therefore the
rendering should not write to the database, which is the case for all
major scene graph implementations.

@@

The following contextual information has to be used in order to render
the view expected by Equalizer:
\begin{description}
\item[Buffer]: The OpenGL read and draw buffer as well as color mask.
  These parameters are influenced by the current eye pass, eye
  seperation and anaglyphic stereo settings.
\item[Viewport]: The two-dimensional pixel viewport restricting the
  rendering area within the channel. For correct operations, both
  \textsf{glViewport} and \textsf{glScissor} have to be used. The pixel
  viewport is influenced by the destination channel's viewport
  definition and compound viewports set for sort-first/2D decompositions.
\item[Frustum]: Frustum parameters as defined by
  \textsf{glFrustum}. Typically the frustum used to set up the OpenGL
  projection matrix. The frustum is influenced by the destination
  channel's view definition, compound viewports, head matrix and the
  current eye pass.
\item[Head Transformation:] A transformation matrix positioning the
  frustum. This is typically an identity matrix and is used for off-axis
  frustra in immersive rendering. It is normally used to set up the
  `view' part of the modelview matrix, before static light sources are
  defined.
\item[Range]: A one-dimensional range with the interval [0..1]. This
  parameter is optional and should be used by the application to render
  only the appropriate subset of its data.
\end{description}

@@

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/DB.pdf}
  {\caption{\small\label{fDB}Destination view of an DB compound}}
\end{wrapfigure}

After the basic view setup, a directional light is configured, and the
model is positioned using the camera parameters from the frame data. The
camera parameters are transported using the the frame data to ensure
that all channels render the frame in the same position.

@@

Finally the model is rendered, if it could be loaded during
node initialization. Otherwise a quad is drawn in its place:

@@

In order to draw the model, a helper class for view frustum culling is
set up using the view frustum from Equalizer and the camera
position. The frustum helper computes the six frustum planes from the
projection and modelView matrices. During rendering, the bounding
spheres of the model are tested against these planes to determine the
visibility with the frustum. Furthermore the render state from the
window, and the database range from the channel is obtained. The render
state manages display list or VBO allocation:

{\footnotesize\begin{lstlisting}
void Channel::_drawModel( const Model* model )
{
    Window*                  window    = static_cast<Window*>( getWindow() );
    mesh::VertexBufferState& state     = window->getState();

@@

\begin{wrapfigure}{r}{.6\textwidth}
  \includegraphics[width=.6\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Rendering Loop}}
\end{wrapfigure}
The model data is spatially organized in an 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree and each
node of the tree gets a database range assigned, that is, the root node
has the range [0, 1], its left child [0, 0.5] and its right child [0.5,
1], and so on for all nodes in the tree. The preprocessed model is saved
in a binary format for accelerating subsequent use.

The rendering loop maintains a list of candidates to render, which
initially contains the root node. Each candidate of this list is tested
for full visiblity against the frustum and range, and rendered if
visible. It is dropped if it is fully invisible or fully out of
range. If it is partially visible or partially in range, the children of
the node are added to the candidate list:

{\footnotesize\begin{lstlisting}
    model->beginRendering( state );

    // start with root node
    vector< const VertexBufferBase* > candidates;
    candidates.push_back( model );

    while( !candidates.empty() )
    {
        const VertexBufferBase* treeNode = candidates.back();
        candidates.pop_back();

        // completely out of range check
        if( treeNode->getRange()[0] >= range.end || 
            treeNode->getRange()[1] < range.start )
            continue;

        // bounding sphere view frustum culling
        switch( culler.testSphere( treeNode->getBoundingSphere() ) )
        {
            case vmml::VISIBILITY_FULL:
                // if fully visible and fully in range, render it
                if( treeNode->getRange()[0] >= range.start && 
                    treeNode->getRange()[1] < range.end )
                {
                    treeNode->render( state );
                    break;
                }
                // partial range, fall through to partial visibility
            case vmml::VISIBILITY_PARTIAL:
            {
                const VertexBufferBase* left  = treeNode->getLeft();
                const VertexBufferBase* right = treeNode->getRight();

                if( !left && !right )
                {
                    if( treeNode->getRange()[0] >= range.start )
                        treeNode->render( state );
                    // else drop, to be drawn by 'previous' channel
                }
                else
                {
                    if( left )
                        candidates.push_back( left );
                    if( right )
                        candidates.push_back( right );
                }
                break;
            }
            case vmml::VISIBILITY_NONE:
                // do nothing
                break;
        }
    }

    model->endRendering( state );
}
\end{lstlisting}}

@@

This section discusses some additional, important features not covered
by the previous \textsf{eqPly} section. Where possible, code examples
from the Equalizer distribution are used.

@@

Event handling requires a lot of flexibility. On one hand, the
implementation differs slightly for each operating and window system due
to conceptual differences in the OS-specific implementation. On the
other hand, each application and widget set has its own model on how
events are to be handled. Therefore, event handling in Equalizer is
customizable in any stage of the processing, to the extreme of making it
possible to disable all event-related code in Equalizer. More
information on event handling can be found on the Equalizer
website\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/eventHandling.html}}.

The default implementation provides a convenient, easily accessible
event framework, while allowing all necessary customizations. It gathers
all events in the application main thread, so that the developer only
has to implement \textsf{Config::processEvent} to update its data based
on the pre-processed, generic keyboard and mouse events. It is very easy
to use and similar to an GLUT-based implementation.

@@

Where possible, events are received and processed by a separate per-node
event thread to allow asynchronous\footnote{with respect to the
  rendering} event handling. Currently an event thread is only used by
the X11/glX window system. WGL receives and processes the events from
the pipe threads which created the windows. AGL receives the events from
application or node main thread. Whenever the term \textbf{event thread}
is used, it refers to the thread receiving the event, i.e., a per-node
thread for glX, the pipe thread for WGL and the main thread for AGL.

@@

During window and pipe initialization the event handling is set up. For
both entities, \textsf{initEvent\-Handler} is called to register the
pipe or window with an event handler. This method may be overwritten to
use a custom event handler, or to not install an event handler to
disable event handling. Likewise, \textsf{exitEventHandler} is called to
de-initialize event handling.

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/eventInit.pdf}
  {\caption{\small\label{fEventInit}Event Handling Initialization}}
\end{wrapfigure}
An event handler consists of two parts: the generic base class providing
the interface and generic functions, and the window-system-specific
part providing the actual implementation.

@@

For the WGL and AGL window system it is required to manually receive and
dispatch (`pump') events. On WGL, this has to happen on each thread with
windows, whereas on AGL it has to happen only on the main thread. By
default, Equalizer pumps these events automatically for the application.

The methods \textsf{Client::useMessagePump} and
\textsf{Pipe::useMessagePump} can be overridden to return \textsf{false}
to disable this behaviour for their respective threads. On non-threaded
pipes, \textsf{Pipe::useMessagePump} is not called.

If the application disables message pumping in Equalizer, it has to make
sure the events are pumped externally, as it often done within external
widget sets, e.g., Trolltech's Qt.

@@

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/eventFilter.pdf}
  {\caption{\small\label{fEventProcessing}Event Processing}}
\end{wrapfigure}
Events are received by an event handler. The event handler finds the
\textsf{eq::Window} associated to the event. It then creates a generic
\textsf{WindowEvent}, which holds important event data in a format
independent of the window system. The original event is attached to the
generic window event.

@@

Events sent using \textsf{Config::sendEvent} are queued in the
application thread. After a frame has been finished,
\textsf{Config::finishFrame} calls \textsf{Config::handleEvents}. The
default implementation of this method provides non-blocking event
processing, that is, it calls \textsf{Config::handleEvent} for each
queued event. By overriding this function, event-driven execution can be
implemented.

Later Equalizer versions will introduce \textsf{Pipe::processEvent} and
\textsf{PipeEvent} to communicate pipe-specific events, e.g, resolution
changes.

@@

\subsection{\label{sCompositing}Image Compositing for Scalable Rendering}

@@

\subsubsection{Parallel Direct Send Compositing}

In order to provide a motivation for the design of the image compositing
API, the parallel compositing algorithm for direct send is introduced
here.

The motivation for direct send is to parallelize the costly reassembly
for database (sort-last) decomposition. With each additional source
channel, the amount of pixel data to be composited grows
linearly. Direct send distributes this workload evenly across all source
channels, and thereby keeps the compositing work per channel constant.

\begin{wrapfigure}{r}{.6\textwidth}
  \includegraphics[width=.6\textwidth]{images/directSend.pdf}
  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
\end{wrapfigure}
Each channel is responsible for the sort-last compositing of one 2D
tile. He receives the framebuffer pixels for his tike from all the other
channels. The size of this tile decreases linearly with the number of
source channels.

After performing the sort-last compositing, the color information is
transferred to the destination channel, similarly to an 2D (sort-first)
compound.

\fig{fDirectSend} illustrates this algorithm for three channels. The
Equalizer website contains a presentation for this
algorithm\footnote{\link{http://www.equalizergraphics.com/documents/EGPGV07.pdf}}.

The following operations have to be possible in order to perform this
algorithm:
\begin{itemize}
\item Selection of color and/or depth readback
\item Restricting the read back to a part of the rendered area
\item Positioning the pixel data correctly on the receiving channels
\end{itemize}

Furthermore it should be possible for the application to implement a
read back of only the relevant region of interest, that is, the 2D area
of the framebuffer actually updated during rendering. This optimization
will be fully supported by later versions of Equalizer.

@@

An \textsf{eq::Frame} references an \textsf{eq::Fra\-me\-Data}. The
frame data is the object connecting output with input frames. Output and
input frames with the same name and from the same compound tree will
reference the same frame data.

The frame data is a holder for images and is used to signal the
availability of the pixel data. An \textsf{eq::Image} holds a
two-dimensional snapshot of the framebuffer and can contain color and/or
depth information.

The frame data transports the inherited range of its compound. The range
is needed to compute the assembly order of multiple input frames, e.g.,
for sorted-blend compositing in volume rendering applications.

Readback and assemble operations on frames and images are designed to be
asynchronous. They have a start and finish method for both readback and
assemble to allow the initiation and synchronization of the operation.
Currently, only synchronous readback and assembly using
\textsf{glReadPixels} and \textsf{glDrawPixels} is implemented in the
respective start method of the image.

The offset of input and output frames characterizes the position of the
frame data with respect to the framebuffer, that is, the
\textbf{window's} lower-left corner. For output frames this is the
position of the channel with respect to the window.

\begin{wrapfigure}{r}{.6\textwidth}
  \includegraphics[width=.6\textwidth]{images/assembly.pdf}
  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
\end{wrapfigure}
For output frames, the frame data's pixel viewport (pvp) is the area of
the frame buffer to read back. It will transport the offset from the
source to the destination channel, that is, the frame data pvp for input
frames position the pixel data on the destination. This has the effect
that a partial framebuffer readback will end up in the same place in the
destination channels.

The image pixel viewport signifies the region of interest read
back. Currently this pvp has no offset and is as big as the frame data
pvp. Only one image will be read back.

\fig{fAssembly} illustrates the relationship between frames, frame data
and images.
